Instructions for developing new pages for the MarineOmics webpage

### 2023-12-15 DRAFT_DNA_methylation.Rmd
Update DRAFT_DNA_methylation.Rmd
added vert methylation distribution image
meth - ref formatting

The purpose of this tutorial is to provide an example of aligning bisulfite-treated and enzymatically converted DNA sequence data to an existing genome, and discuss needs for potential downstream applications. The principle behind creating bisulfite-treated DNA libraries is that when cytosines are not methylated, then they will be converted to uracil, which ultimately result in thymine nucleotides in the sequence data. [@https://doi.org/10.1111/1755-0998.13542](#refs) offers a comparison of three methods for quantifying DNA methylation at single base-pair resolution using whole genome bisulfite sequencing (WGBS), reduced representation bisulfite sequencing (RRBS), and methyl-CpG binding domain bisulfite sequencing (MBDBS). When aligning bisulfite-converted data to a reference genome the percent methylation at a given CpG loci is determined by examining the ratio of thymines to cytosines. For example if there is a given cytosine adjacent to a guanine (ie CpG locus) with 30% of the reads containing thymine, one would consider this cytosine loci (CpG) to be 70% methylated. Since bisulfite treatment can damage DNA, newer library preparation and sequencing methods (ex. [NEBNext® Enzymatic Methyl-seq Kit; EM-seq](https://www.neb.com/en-us/products/e7120-nebnext-enzymatic-methyl-seq-kit#Protocols,%20Manuals%20&%20Usage) are starting to gain popularity. As EM-Seq produces loci-level data, alignment of EM-Seq data is similar to WGBS. Also, technologies like Nanopore and PacBio sequencing can directly detect DNA methylation without bisulfite conversion. As DNA passes through a nanopore or is sequenced by single-molecule real-time (SMRT) technology, changes in electrical current or fluorescence patterns can indicate the presence of methylated bases.

@@

For WGBS and MBDBS data, it is general practice to hard-trim 10 bp from the beginning and end of each read. [RRBS data often needs different trimming specifications](https://rawgit.com/FelixKrueger/Bismark/master/Docs/Bismark_User_Guide.html#ix-notes-about-different-library-types-and-commercial-kits).

@@

-   Gene-level methylation analysis: see [@. It is worth considering the utility of converting loci-level resolution data to gene-level resolution data.

-   Differential loci methylation and/or region analyses. example: [@10.1186/s12864-022-08781-5](#refs). methylKit is a popular package to conduct this analysis differential methylation anlaysis
-   Feature overlap. see [@10.1186/s12864-022-08781-5](#refs) Bedtools for quantitative assessment and IGV for visual assessment are valuable for this.

>>>>

The purpose of this tutorial is to provide an example of aligning bisulfite-treated and enzymatically converted DNA sequence data to an existing genome, and discuss needs for potential downstream applications. The principle behind creating bisulfite-treated DNA libraries is that when cytosines are not methylated, then they will be converted to uracil, which ultimately result in thymine nucleotides in the sequence data. [@https://doi.org/10.1111/1755-0998.13542](#refs) offers a comparison of three methods for quantifying DNA methylation at single base-pair resolution using whole genome bisulfite sequencing (WGBS), reduced representation bisulfite sequencing (RRBS), and methyl-CpG binding domain bisulfite sequencing (MBDBS). When aligning bisulfite-converted data to a reference genome the percent methylation at a given CpG loci is determined by examining the ratio of thymines to cytosines. For example if there is a given cytosine adjacent to a guanine (ie CpG locus) with 30% of the reads containing thymine, one would consider this cytosine loci (CpG) to be 70% methylated. Since bisulfite treatment can damage DNA, newer library preparation and sequencing methods (ex. [NEBNext® Enzymatic Methyl-seq Kit; EM-seq](https://www.neb.com/en-us/products/e7120-nebnext-enzymatic-methyl-seq-kit#Protocols,%20Manuals%20&%20Usage) are starting to gain popularity. As EM-Seq produces loci-level data, alignment of EM-Seq data is similar to WGBS. Also, technologies like Nanopore and PacBio sequencing can directly detect DNA methylation without bisulfite conversion. See [@10.1093/g3journal/jkab148](#refs) as example of using Nanopore sequencing. As DNA passes through a nanopore or is sequenced by single-molecule real-time (SMRT) technology, changes in electrical current or fluorescence patterns can indicate the presence of methylated bases.

@@

For WGBS and MBDBS data, it is general practice to hard-trim 10 bp from the beginning and end of each read. RRBS data often needs different trimming specifications [see Manual for details](https://felixkrueger.github.io/Bismark/)

@@

-   Gene-level methylation analysis: example [@10.1101/2022.06.24.497506](#refs). It is worth considering the utility of converting loci-level resolution data to gene-level resolution data.

-   Differential loci methylation and/or region analyses. example: [@10.1186/s12864-022-08781-5](#refs). methylKit is a popular package to conduct this analysis differential methylation anlaysis

-   Feature overlap. see [@10.1186/s12864-022-08781-5](#refs) Bedtools for quantitative assessment and IGV for visual assessment are valuable for this.
